noam_data_exploration_v1

### 2024-08-04 noam_data_exploration_v3.Rmd
Updated for use with the corrected colnames file

```{r import and format data}
# starting with all data now because of 0 algae transects/quadrats
data <- read.csv(here::here("data", "allbenthicdata_cleaned.csv"), stringsAsFactors = F)

not_algae <- c("Ascidian", "Bare.Space", "Coral", "Coral.Rubble", "Corallimorpharia", "Heteractis.sp.", "Millepora.platyphylla", "No.data", "Sand", "Sacrophyton.sp.", "Shell.Debris", "Soft.Coral", "Sponge", "Tridacna.sp.", "Zooxanthellae", "row_sums")

@@

# adding island side - adapted from Julianna's code
data <- data %>% mutate(Island_side = case_when(Site == "LTER 1" ~ "North",
                                 Site == "LTER 2" ~ "North",
                                 Site == "LTER 3" ~ "East",
                                 Site == "LTER 4" ~ "East",
                                 Site == "LTER 5" ~ "West",
                                 Site == "LTER 6" ~ "West"))
data$Island_side_habitat <- paste(data$Island_side, data$Habitat)

@@

meta_cols <- c(colnames(data[1:7] ), "Island_side", "Island_side_habitat", "Stratum")
species_cols <- colnames(data[8:77])

#pivot long 
long_data <- pivot_longer(
  data = data,
  cols = 8:77,
  names_to = "Taxa",
  values_to = "Percent_cover")

 transect_level <- long_data %>%
   dplyr::group_by(Year, Site, Habitat, Transect, Taxa) %>%
   dplyr::summarise(Mean_cover = mean(Percent_cover), 
             SE = sd(Percent_cover)/sqrt(n()))

# site_level <- transect_level %>%
#   dplyr::group_by(Year, Site, Habitat, Taxa) %>%
#   dplyr::summarise(Mean_site_cover = mean(Mean_cover), 
#             Site_se = sd(Mean_cover)/sqrt(n()))
```

@@

wide_transect_level <- data %>%
  dplyr::group_by(Year, Island_side, Island_side_habitat, Site, Habitat, Site_habitat, Transect) %>%
  dplyr::summarize(across(all_of(species_cols), mean, na.rm = TRUE))

alpha_diversity <-
  data.frame(
    Year = as.factor(wide_transect_level$Year),
    Island_side = as.factor(wide_transect_level$Island_side),
    Island_side_habitat = as.factor(wide_transect_level$Island_side_habitat),
    Site = as.factor(wide_transect_level$Site),
    Habitat = as.factor(wide_transect_level$Habitat),
    Site_habitat = as.factor(wide_transect_level$Site_habitat),
    Transect = as.factor(wide_transect_level$Transect),
    Site_habitat_transect = as.factor(paste(wide_transect_level$Site_habitat, wide_transect_level$Transect)),
    Shannon = diversity(wide_transect_level[ , -which(names(wide_transect_level) %in% meta_cols)], "shannon"),
    Richness = rowSums(wide_transect_level[ , -which(names(wide_transect_level) %in% meta_cols)] > 0)
  )

alpha_diversity_means <- alpha_diversity %>%
  dplyr::group_by(Year, Island_side, Island_side_habitat, Site, Habitat, Site_habitat) %>%
  dplyr::summarise(Richness_mean = mean(Richness), 
            Richness_se = sd(Richness)/sqrt(n()),
            Shannon_mean = mean(Shannon), 
            Shannon_se = sd(Shannon)/sqrt(n()))

site_colours <- c("LTER 1" = "cornflowerblue",
               "LTER 2" = "darkblue",
               "LTER 3" = "yellowgreen",
               "LTER 4" = "forestgreen",
               "LTER 5" = "mediumorchid",
               "LTER 6" = "orange")
island_side_colours <- c("North" = "darkblue",
                         "East" = "mediumorchid",
                         "West" = "orange")
habitat_linetypes <- c("Fringing" = "solid",
                       "Backreef" = "longdash",
                       "Outer 10" = "twodash",
                       "Outer 17" = "dotted")
habitat_shapes <- c("Fringing" = 16,
                     "Backreef" = 17,
                     "Outer 10" = 1,
                     "Outer 17" = 2)

(richness_plot <-
  ggplot(data = alpha_diversity_means) +
  geom_point(aes(x = Year, y = Richness_mean, colour = Island_side, shape = Habitat, group = Island_side_habitat), size = 8) +
  geom_line(aes(x = Year, y = Richness_mean, colour = Island_side, group = Island_side_habitat), size = 1.2) +
  geom_errorbar(aes(x = Year, ymin = Richness_mean - Richness_se , ymax = Richness_mean + Richness_se, colour = Island_side,
                    group = Island_side_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = island_side_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Habitat)+
  labs(title = "a. Richness\n", y = "Richness\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

(shannon_plot <-
  ggplot(data = alpha_diversity_means) +
  geom_point(aes(x = Year, y = Shannon_mean, colour = Island_side, shape = Habitat, group = Island_side_habitat), size = 8) +
  geom_line(aes(x = Year, y = Shannon_mean, colour = Island_side, group = Island_side_habitat), size = 1.2) +
  geom_errorbar(aes(x = Year, ymin = Shannon_mean - Shannon_se , ymax = Shannon_mean + Shannon_se, colour = Island_side,
                    group = Island_side_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = island_side_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Habitat)+
  labs(title = "b. Shannon\n", y = "shannon\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

@@

```{r alpha diversity quadrat level, fig.height = 15, fig.width=20}
alpha_diversity_quad <-
  data.frame(
    Year = as.numeric(data$Year),
    Location = as.factor(data$Location),
    Island_side = as.factor(data$Island_side),
    Island_side_habitat = as.factor(data$Island_side_habitat),
    Site = as.factor(data$Site),
    Habitat = as.factor(data$Habitat),
    Site_habitat = as.factor(data$Site_habitat),
    Transect = as.factor(data$Transect),
    Site_habitat_transect = as.factor(paste(data$Site_habitat, data$Transect)),
    Shannon = diversity(data[ , -which(names(data) %in% meta_cols)], "shannon"),
    Richness = rowSums(data[ , -which(names(data) %in% meta_cols)] > 0)
  )

alpha_diversity_quad_means <- alpha_diversity_quad %>%
  dplyr::group_by(Year, Island_side, Island_side_habitat, Site, Habitat, Site_habitat) %>%
  dplyr::summarise(Richness_mean = mean(Richness), 
            Richness_se = sd(Richness)/sqrt(n()),
            Shannon_mean = mean(Shannon), 
            Shannon_se = sd(Shannon)/sqrt(n()))

(richness_quad_plot <-
  ggplot(data = alpha_diversity_quad_means) +
  geom_point(aes(x = Year, y = Richness_mean, colour = Site, shape = Habitat, group = Site_habitat), size = 8) +
  geom_line(aes(x = Year, y = Richness_mean, colour = Site, group = Site_habitat), size = 1.2) +
  geom_errorbar(aes(x = Year, ymin = Richness_mean - Richness_se , ymax = Richness_mean + Richness_se, colour = Site,
                    group = Site_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = site_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Habitat)+
  labs(title = "a. Richness quad level\n", y = "Richness\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

(shannon_quad_plot <-
  ggplot(data = alpha_diversity_quad_means) +
  geom_point(aes(x = Year, y = Shannon_mean, colour = Island_side, shape = Habitat, group = Island_side_habitat), size = 8) +
  geom_line(aes(x = Year, y = Shannon_mean, colour = Island_side, group = Island_side_habitat), size = 1.2) +
  geom_errorbar(aes(x = Year, ymin = Shannon_mean - Shannon_se , ymax = Shannon_mean + Shannon_se, colour = Island_side,
                    group = Island_side_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = island_side_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Habitat)+
  labs(title = "b. Shannon quad level\n", y = "shannon\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

```
```{r compare quad and transect level plots, fig.height = 30, fig.width= 40}
ggarrange(richness_plot + rremove("x.text"), richness_quad_plot + rremove("x.text"),
          shannon_plot,shannon_quad_plot,
          common.legend = T,
          nrow = 2, ncol = 2)

```
```{r gam approach}
# NOTE: Not sure where I'm going with this yet
# Idea is to look at smoother estimates for each time series to get an idea for the "shape" of the time series
# Not sure if worthwhile, might skip

# The below takes 17 years to run in and of itself
# gam_model <- gam(Richness ~ s(Year) + s(Year, by = Habitat) + 
#                  s(Island_side, bs = "re") + s(Transect, bs = "re") + s(Location, bs = "re"),
#                  data = alpha_diversity_quad, family = poisson())

gam_columns <- c("Location", "Intercept", "edf") # names of columns for dataframe
gam_coefficients <- # new dataframe name - same workflow as above
  data.frame(matrix(nrow = length(unique(alpha_diversity_quad$Location)),
                    ncol = length(gam_columns))
             )
colnames(gam_coefficients) <- gam_columns # change column names to what was specified
gam_coefficients$Location <- levels(alpha_diversity_quad$Location) # populate Location column with each quad/time series

# same workflow as *below (I did gls first originally)
for(i in 1:length(unique(alpha_diversity_quad$Location)) ){ # for each  quadrat
  # create a new dataframe for the quadrat of the current iteration
  newData <- subset(alpha_diversity_quad, Location == levels(alpha_diversity_quad$Location)[i])

  # fit gam for current time series
  loop_mod <- gamm(Richness ~ s(Year), data = newData, correlation = corCAR1(form = ~ Year), method = "REML", family = poisson)
  loop_sum <- summary(loop_mod$gam) # store summary data
  # populate appropriate row and column with the relevant coeffcients from gls
  gam_coefficients[i, 2] <- unname(loop_sum$p.coeff)
  gam_coefficients[i, 3] <- unname(loop_sum$edf)

@@

hist(gam_coefficients$Intercept)
hist(gam_coefficients$edf) # generally around 1 except for an outlier
# LTER 5 Fringing Reef Algae Transect 3 Quad 6 - a quadrat that has almost no algae the entire time series
# issues with quadrat - smaller sample = more 0s, makes modeling trickier
hist(subset(gam_coefficients, Location != "LTER 5 Fringing Reef Algae Transect 3 Quad 6")$edf)
sum(gam_coefficients$edf > 1.01) # all but 18/1200 quadrats can be fit by a smoother with edf < 1.01
# trends more straightforward than I anticipated?
# argument for just looking at mean richness and slope?
```

@@

columns <- c("Location", "Intercept", "Slope") # names of columns for dataframe
gls_coefficients <- # new dataframe name
  data.frame(matrix(nrow = length(unique(alpha_diversity_quad$Location)), # 1200 quadrats = 1200 different time series
                    ncol = length(columns)) # specified to make sure number of columns doesn't exceed what was specified
             )
colnames(gls_coefficients) <- columns # change column names to what was specified
gls_coefficients$Location <- levels(alpha_diversity_quad$Location) # populate Location column with each quad/time series

for(i in 1:length(unique(alpha_diversity_quad$Location)) ){ # for each  quadrat
  # create a new dataframe for the quadrat of the current iteration
  newData <- subset(alpha_diversity_quad, Location == levels(alpha_diversity_quad$Location)[i])

  # fit gls for current time series
  loop_mod <- gls(Richness ~ Year, correlation =corAR1(form=~Year), data = newData)

@@

# Start adding back in relevant metadata - probably a more elegant way to do this!!!
habitat_vector <- c(rep("Backreef", 50), rep("Fringing", 50), rep("Outer 10", 50), rep("Outer 17", 50))
transect_sequence <- c(rep(1,10),
                       rep(2,10),
                       rep(3,10),
                       rep(4,10),
                       rep(5,10))
extra_meta <- 
  data.frame(Site = as.factor(c(rep("LTER 1", 200),
                                rep("LTER 2", 200),
                                rep("LTER 3", 200),
                                rep("LTER 4", 200),
                                rep("LTER 5", 200),
                                rep("LTER 6", 200)) ),
             Island_side = as.factor(c(rep("North", 400),
                                     rep("East", 400),
                                     rep("West", 400))
                                     ),
             Habitat = as.factor(rep(habitat_vector, 6 )),
             Transect = as.factor(rep(transect_sequence, 240)),
             Location = levels(alpha_diversity_quad$Location) )

gls_coefficients_merged <- merge(gls_coefficients, extra_meta, by = "Location")

>>>>

```{r import and format data}
# starting with all data now because of 0 algae transects/quadrats
data <- read.csv(here::here("data", "allbenthicdata_cleaned_colnames.csv"), stringsAsFactors = F)

not_algae <- c("ascidian", "bare_space", "coral", "coral_rubble", "corallimorpharia", "heteractis_sp", "millepora_platyphylla", "no_data", "sand", "sacrophyton_sp", "shell_debris", "soft_coral", "sponge", "tridacna_sp", "zooxanthellae", "row_sums")

@@

# adding island side - adapted from Julianna's code
data <- data %>% mutate(island_side = case_when(site == "lter_1" ~ "north",
                                 site == "lter_2" ~ "north",
                                 site == "lter_3" ~ "east",
                                 site == "lter_4" ~ "east",
                                 site == "lter_5" ~ "west",
                                 site == "lter_6" ~ "west"))
data$island_side_habitat <- paste(data$island_side, data$habitat)

@@

meta_cols <- c(colnames(data[1:7] ), "island_side", "island_side_habitat", "stratum")
species_cols <- colnames(data)[!colnames(data) %in% meta_cols]

#pivot long 
long_data <- pivot_longer(
  data = data,
  cols = 8:77,
  names_to = "taxa",
  values_to = "percent_cover")

 transect_level <- long_data %>%
   dplyr::group_by(year, island_side, site, habitat, transect, taxa) %>%
   dplyr::summarise(mean_cover = mean(percent_cover), 
             SE = sd(percent_cover)/sqrt(n()))

site_level <- transect_level %>%
  dplyr::group_by(year, island_side, site, habitat, taxa) %>%
  dplyr::summarise(mean_site_cover = mean(mean_cover), 
            site_se = sd(mean_cover)/sqrt(n()))
```

@@

wide_transect_level <- data %>%
  dplyr::group_by(year, island_side, island_side_habitat, site, habitat, site_habitat, transect) %>%
  dplyr::summarize(across(all_of(species_cols), mean, na.rm = TRUE))

alpha_diversity <-
  data.frame(
    year = as.factor(wide_transect_level$year),
    island_side = as.factor(wide_transect_level$island_side),
    island_side_habitat = as.factor(wide_transect_level$island_side_habitat),
    site = as.factor(wide_transect_level$site),
    habitat = as.factor(wide_transect_level$habitat),
    site_habitat = as.factor(wide_transect_level$site_habitat),
    transect = as.factor(wide_transect_level$transect),
    site_habitat_transect = as.factor(paste(wide_transect_level$site_habitat, wide_transect_level$transect)),
    shannon = diversity(wide_transect_level[ , -which(names(wide_transect_level) %in% meta_cols)], "shannon"),
    richness = rowSums(wide_transect_level[ , -which(names(wide_transect_level) %in% meta_cols)] > 0)
  )

alpha_diversity_means <- alpha_diversity %>%
  dplyr::group_by(year, island_side, island_side_habitat, site, habitat, site_habitat) %>%
  dplyr::summarise(richness_mean = mean(richness), 
            richness_se = sd(richness)/sqrt(n()),
            shannon_mean = mean(shannon), 
            shannon_se = sd(shannon)/sqrt(n()))

#CALLIE ADDED
#if you need to export this:
#alpha_export <- left_join(alpha_diversity, alpha_diversity_means)
#write_csv(alpha_export, here("data", "algae_alpha_diversity.csv"))

site_colours <- c("lter_1" = "cornflowerblue",
                  "lter_2" = "darkblue",
                  "lter_3" = "yellowgreen",
                  "lter_4" = "forestgreen",
                  "lter_5" = "mediumorchid",
                  "lter_6" = "orange")
island_side_colours <- c("north" = "darkblue",
                         "east" = "mediumorchid",
                         "west" = "orange")
habitat_linetypes <- c("fringing" = "solid",
                       "backreef" = "longdash",
                       "outer_10" = "twodash",
                       "outer_17" = "dotted")
habitat_shapes <- c("fringing" = 16,
                     "backreef" = 17,
                     "outer_10" = 1,
                     "outer_17" = 2)

(richness_plot <-
  ggplot(data = alpha_diversity_means) +
  geom_point(aes(x = year, y = richness_mean, colour = island_side, shape = habitat, group = island_side_habitat), size = 8) +
  geom_line(aes(x = year, y = richness_mean, colour = island_side, group = island_side_habitat), size = 1.2) +
  geom_errorbar(aes(x = year, ymin = richness_mean - richness_se , ymax = richness_mean + richness_se, colour = island_side,
                    group = island_side_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = island_side_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~habitat)+
  labs(title = "a. richness\n", y = "richness\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

(shannon_plot <-
  ggplot(data = alpha_diversity_means) +
  geom_point(aes(x = year, y = shannon_mean, colour = island_side, shape = habitat, group = island_side_habitat), size = 8) +
  geom_line(aes(x = year, y = shannon_mean, colour = island_side, group = island_side_habitat), size = 1.2) +
  geom_errorbar(aes(x = year, ymin = shannon_mean - shannon_se , ymax = shannon_mean + shannon_se, colour = island_side,
                    group = island_side_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = island_side_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~habitat)+
  labs(title = "b. shannon\n", y = "shannon\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

@@

```{r alpha diversity quadrat level, fig.height = 15, fig.width=20}
alpha_diversity_quad <-
  data.frame(
    year = as.numeric(data$year),
    location = as.factor(data$location),
    island_side = as.factor(data$island_side),
    island_side_habitat = as.factor(data$island_side_habitat),
    site = as.factor(data$site),
    habitat = as.factor(data$habitat),
    site_habitat = as.factor(data$site_habitat),
    transect = as.factor(data$transect),
    site_habitat_transect = as.factor(paste(data$site_habitat, data$transect)),
    shannon = diversity(data[ , -which(names(data) %in% meta_cols)], "shannon"),
    richness = rowSums(data[ , -which(names(data) %in% meta_cols)] > 0)
  )

alpha_diversity_quad_means <- alpha_diversity_quad %>%
  dplyr::group_by(year, island_side, island_side_habitat, site, habitat, site_habitat) %>%
  dplyr::summarise(richness_mean = mean(richness), 
            richness_se = sd(richness)/sqrt(n()),
            shannon_mean = mean(shannon), 
            shannon_se = sd(shannon)/sqrt(n()))

(richness_quad_plot <-
  ggplot(data = alpha_diversity_quad_means) +
  geom_point(aes(x = year, y = richness_mean, colour = site, shape = habitat, group = site_habitat), size = 8) +
  geom_line(aes(x = year, y = richness_mean, colour = site, group = site_habitat), size = 1.2) +
  geom_errorbar(aes(x = year, ymin = richness_mean - richness_se , ymax = richness_mean + richness_se, colour = site,
                    group = site_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = site_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~habitat)+
  labs(title = "a. richness quad level\n", y = "richness\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

(shannon_quad_plot <-
  ggplot(data = alpha_diversity_quad_means) +
  geom_point(aes(x = year, y = shannon_mean, colour = island_side, shape = habitat, group = island_side_habitat), size = 8) +
  geom_line(aes(x = year, y = shannon_mean, colour = island_side, group = island_side_habitat), size = 1.2) +
  geom_errorbar(aes(x = year, ymin = shannon_mean - shannon_se , ymax = shannon_mean + shannon_se, colour = island_side,
                    group = island_side_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = island_side_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~habitat)+
  labs(title = "b. shannon quad level\n", y = "shannon\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

```{r compare quad and transect level plots, fig.height = 30, fig.width= 40}
ggarrange(richness_plot + rremove("x.text"), richness_quad_plot + rremove("x.text"),
          shannon_plot,shannon_quad_plot,
          common.legend = T,
          nrow = 2, ncol = 2)

```{r gam approach}
# NOTE: Not sure where I'm going with this yet
# Idea is to look at smoother estimates for each time series to get an idea for the "shape" of the time series
# Not sure if worthwhile, might skip

# The below takes 17 years to run in and of itself
# gam_model <- gam(richness ~ s(year) + s(year, by = habitat) + 
#                  s(island_side, bs = "re") + s(transect, bs = "re") + s(location, bs = "re"),
#                  data = alpha_diversity_quad, family = poisson())

gam_columns <- c("location", "Intercept", "edf") # names of columns for dataframe
gam_coefficients <- # new dataframe name - same workflow as above
  data.frame(matrix(nrow = length(unique(alpha_diversity_quad$location)),
                    ncol = length(gam_columns))
             )
colnames(gam_coefficients) <- gam_columns # change column names to what was specified
gam_coefficients$location <- levels(alpha_diversity_quad$location) # populate location column with each quad/time series

# same workflow as *below (I did gls first originally)
for(i in 1:length(unique(alpha_diversity_quad$location)) ){ # for each  quadrat
  # create a new dataframe for the quadrat of the current iteration
  newData <- subset(alpha_diversity_quad, location == levels(alpha_diversity_quad$location)[i])

  # fit gam for current time series
  loop_mod <- gamm(richness ~ s(year), data = newData, correlation = corCAR1(form = ~ year), method = "REML", family = poisson)
  loop_sum <- summary(loop_mod$gam) # store summary data
  # populate appropriate row and column with the relevant coeffcients from gls
  gam_coefficients[i, 2] <- unname(loop_sum$p.coeff)
  gam_coefficients[i, 3] <- unname(loop_sum$edf)

@@

hist(gam_coefficients$Intercept)
hist(gam_coefficients$edf) # generally around 1 except for an outlier
# lter_5 fringing Reef Algae transect 3 Quad 6 - a quadrat that has almost no algae the entire time series
# issues with quadrat - smaller sample = more 0s, makes modeling trickier
hist(subset(gam_coefficients, location != "lter_5 fringing Reef Algae transect 3 Quad 6")$edf)
sum(gam_coefficients$edf > 1.01) # all but 18/1200 quadrats can be fit by a smoother with edf < 1.01
# trends more straightforward than I anticipated?
# argument for just looking at mean richness and slope?
```

@@

columns <- c("location", "Intercept", "Slope") # names of columns for dataframe
gls_coefficients <- # new dataframe name
  data.frame(matrix(nrow = length(unique(alpha_diversity_quad$location)), # 1200 quadrats = 1200 different time series
                    ncol = length(columns)) # specified to make sure number of columns doesn't exceed what was specified
             )
colnames(gls_coefficients) <- columns # change column names to what was specified
gls_coefficients$location <- levels(alpha_diversity_quad$location) # populate location column with each quad/time series

for(i in 1:length(unique(alpha_diversity_quad$location)) ){ # for each  quadrat
  # create a new dataframe for the quadrat of the current iteration
  newData <- subset(alpha_diversity_quad, location == levels(alpha_diversity_quad$location)[i])

  # fit gls for current time series
  loop_mod <- gls(richness ~ year, correlation =corAR1(form=~year), data = newData)

@@

# Start adding back in relevant metadata - probably a more elegant way to do this!!!
habitat_vector <- c(rep("backreef", 50), rep("fringing", 50), rep("outer_10", 50), rep("outer_17", 50))
transect_sequence <- c(rep(1,10),
                       rep(2,10),
                       rep(3,10),
                       rep(4,10),
                       rep(5,10))
extra_meta <- 
  data.frame(site = as.factor(c(rep("lter_1", 200),
                                rep("lter_2", 200),
                                rep("lter_3", 200),
                                rep("lter_4", 200),
                                rep("lter_5", 200),
                                rep("lter_6", 200)) ),
             island_side = as.factor(c(rep("north", 400),
                                     rep("east", 400),
                                     rep("west", 400))
                                     ),
             habitat = as.factor(rep(habitat_vector, 6 )),
             transect = as.factor(rep(transect_sequence, 240)),
             location = levels(alpha_diversity_quad$location) )

gls_coefficients_merged <- merge(gls_coefficients, extra_meta, by = "location")

slope_mod <- glmmTMB(Slope ~ habitat*site + (1|site/transect), data = gls_coefficients_merged)
# model convergence problem
summary(slope_mod) # pretty low variance in random effects
plot(gls_res <- simulateResiduals(slope_mod)) # there are issues, but I can't figure out what they are...
# I think this is due to 1200 obs in model
hist(residuals(slope_mod))
plot(predict(slope_mod) ~ residuals(slope_mod))
Anova(slope_mod)
# Response: Slope
#                 Chisq Df Pr(>Chisq)    
# habitat      1397.999  3  < 2.2e-16 ***
# site           39.095  5  2.273e-07 ***
# habitat:site 1738.155 15  < 2.2e-16 ***
r.squaredGLMM(slope_mod) # 0.22, 0.24
plot(allEffects(slope_mod))
# backreef: richness increases on east and north shore, but relatively constant west side
# Fringe: richness decreasing on east and west side, but relatvely constant on north shore
# outer_10: richness neutral/decreasing on east side, but neutral/increasing on north and west side
# outer_17: richness decreasing north/east side, but neutral west side
# Worth noting trends are slight - arguable whether they are biologically meaningful

slope_emm <- emmeans(slope_mod, ~ habitat*site) # create emmeans object
emmeans::contrast(slope_emm, "consec", simple = "each", combine = TRUE, adjust = "bonferroni")
# Note: not sure I want to include this
emmeans(slope_mod, pairwise ~ site | habitat)
```
